---
title: "fire_analysis"
author: "Anna J Holmquist"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'F:/Research/rprojects/fire_project/') 
# How do we make this relative for reproducibility?
```

```{r libraries}

library(tidyverse)
library(reshape2)
library(ape)
library(vegan)
library(BAT)
library(venn)
library(ggpubr)

```

```{r data}
getwd()
# Produced from cleanup
df_97 <- read.csv("mco_97_wsites.csv")
sites <- read.csv("site_data.csv")

```

Taxonomy summary
```{r taxonomy summary}

df_97 %>%
  summarise(fam = n_distinct(family),
            gen = n_distinct(genus),
            sp = n_distinct(species),
            order = n_distinct(order))

# Venn
families <- list(
  Burned = unique(df_97$family[df_97$status == "Burn"
                               & !is.na(df_97$family)]),
  Unburned = unique(df_97$family[df_97$status == "Unburn"
                               & !is.na(df_97$family)])
  )

venn(families, opacity = 0.4, zcolor = c("#d0681e","#1e74d0"))

# Not included currently
genera <- list(
  Burned = unique(df_97$genus[df_97$status == "Burn"
                               & !is.na(df_97$family)]),
  Unburned = unique(df_97$genus[df_97$status == "Unburn"
                               & !is.na(df_97$family)])
  )

venn(genera, opacity = 0.4, zcolor = c("#d0681e","#1e74d0"))

# Bars
library(viridis)
colors <- rocket(n = 11, begin = 1, end = 0)

order_bar <- 
  df_97 %>%
  filter(!is.na(order)) %>%
  group_by(year, order, status) %>%
  summarise(n = n_distinct(otu)) %>% 
  ungroup() %>%
  arrange(year, desc(n)) %>%
  mutate(order = factor(order, levels = unique(order))) %>%
  mutate(year = ifelse(year == "pan2020", "Fall 2020", "Spring 2021")) %>%
  mutate(status = ifelse(status == "Burn", "Burned", "Unburned")) %>%
  rename(Order = order) %>%
  ggplot(aes(x = year, y = n, fill = Order)) +
  geom_bar(stat = "identity",  color = "black", alpha = 0.8) + 
  facet_grid(~status) +
  theme_minimal() +
  #theme(axis.text.x = element_text(angle = 45)) +
  scale_fill_manual(values = colors) +
  ylab("Number of OTUs") +
  xlab("")

ggsave("family_bar.png", width = 6, height = 4, units = "in")

```

Site summary
```{r habitat}
df_filtered <-
  df_97 %>%
  filter(reserve != "QR" & !is.na(habitat) 
         & habitat %in% c("Oak Woodland", "Scrub", "Grassland")) %>%
  ungroup()

```

Hypothesis 1: Species richness will not be different between burned and unburned sites
```{r richness, alpha diversity}

# Create dataframe with reads by OTU at each site and year
otu_counts <- 
  df_filtered %>%
  group_by(sample_name, otu) %>%
  # Remove multiple rows
  filter(row_number() == 1) %>%
  # Use reads across all replicates
  summarise(otu_count = sum(rep_combined)) %>%
  ungroup()

# Create Hellinger, incidence matrix
otu_mat <-
  otu_counts %>%
  select(sample_name, otu, otu_count) %>%
  acast(sample_name ~ otu, fill = 0)

hell_mat <- decostand(otu_mat, method = "hellinger")
inc_mat <- otu_mat
inc_mat[inc_mat > 0] <- 1

# Calculate alpha diversity
diversity_shannon <- as.data.frame(vegan::diversity(hell_mat, index = "shannon")) %>%
    rownames_to_column(var = "id")
colnames(diversity_shannon) <- c("sample_name", "div")

hist_shannon <- 
  diversity_shannon %>%
  ggplot(aes(div)) + 
  geom_histogram(bins = 10, color = "black", alpha = 0.5) +
  theme(title = "Distribution - Shannon index") +
  xlab("Shannon index") +
  ylab("Count") +
  theme_minimal()

diversity_simpson <- as.data.frame(vegan::diversity(hell_mat, index = "simpson")) %>%
    rownames_to_column(var = "id")
colnames(diversity_simpson) <- c("sample_name", "div")

hist_simp <- 
  diversity_simpson %>%
  ggplot(aes(div)) + 
  geom_histogram(bins = 10, color = "black", alpha = 0.5) +
  theme(title = "Distribution - Simpson index") +
  xlab("Simpson index") +
  ylab("Count") +
  theme_minimal()

diversity_rich <- as.data.frame(BAT::alpha(hell_mat)) %>%
  rownames_to_column(var = "id")
colnames(diversity_rich) <- c("sample_name", "div")

hist_rich <- 
  diversity_rich %>%
  ggplot(aes(div)) + 
  geom_histogram(bins = 10, color = "black", alpha = 0.5) +
  theme(title = "Richness") +
  xlab("Richness") +
  ylab("Count") +
  theme_minimal()

hist_alpha <- ggarrange(hist_shannon, hist_simp, hist_rich, ncol = 3)

# Join diversity with site data
diversity_rich <- 
  diversity_rich %>%
  left_join(sites, by = "sample_name")

diversity_shannon <- 
  diversity_shannon %>%
  left_join(sites, by = "sample_name")

diversity_simpson <- 
  diversity_simpson %>%
  left_join(sites, by = "sample_name")

# Test variance
var.test(diversity_rich$div[diversity_rich$status == "Burn"], 
         diversity_rich$div[diversity_rich$status == "Unburn"]) 

var.test(diversity_shannon$div[diversity_shannon$status == "Burn"], 
         diversity_shannon$div[diversity_shannon$status == "Unburn"])

var.test(diversity_simpson$div[diversity_simpson$status == "Burn"], 
         diversity_simpson$div[diversity_simpson$status == "Unburn"]) 

# T-test
t_status1 <- 
  wilcox.test(diversity_rich$div[diversity_rich$status == "Burn"], 
         diversity_rich$div[diversity_rich$status == "Unburn"])
rich_p <- t_status1$p.value

t_status2 <- 
  wilcox.test(diversity_shannon$div[diversity_shannon$status == "Burn"], 
         diversity_shannon$div[diversity_shannon$status == "Unburn"])
shan_p <- t_status2$p.value

t_status3 <- 
  wilcox.test(diversity_simpson$div[diversity_simpson$status == "Burn"], 
         diversity_simpson$div[diversity_simpson$status == "Unburn"])

simp_p <- t_status3$p.value

# Mood's Median
library(coin)

median_test(div ~ as.factor(status), diversity_rich)
median_test(div ~ as.factor(status), diversity_shannon)
median_test(div ~ as.factor(status), diversity_simpson)


# Plots c("#d0681e","#1e74d0")
p1 <- ggplot(diversity_rich, aes(status, div)) +
  geom_violin(aes(fill = status), alpha = 0.6) +
  geom_jitter(alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 24, size = 2, 
               color = "black", fill = "white" ) +
  annotate(geom="text", 
           label = paste0("p-value: ", round(rich_p, digits = 7)), 
           x = 2, y = 32) +
  scale_fill_manual(values = c("#d0681e","#1e74d0")) +
  ylab("Richness") +
  xlab("") +
  theme_minimal()

p2 <- ggplot(diversity_shannon, aes(status, div)) +
  geom_violin(aes(fill = status), alpha = 0.6) +
  geom_jitter(alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 24, size = 2, 
               color = "black", fill = "white" ) +
  annotate(geom="text", 
           label = paste0("p-value: ", round(shan_p, digits = 3)), 
           x = 2, y = 3.5) +
  scale_fill_manual(values = c("#d0681e","#1e74d0")) +
  ylab("Shannon Diversity") +
  xlab("") +
  theme_minimal()

p3 <- ggplot(diversity_simpson, aes(status, div)) +
  geom_violin(aes(fill = status), alpha = 0.6) +
  geom_jitter(alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 24, size = 2, 
               color = "black", fill = "white" ) +
  annotate(geom="text", 
           label = paste0("p-value: ", round(simp_p, digits = 4)), 
           x = 1.8, y = 1) + 
  scale_fill_manual(values = c("#d0681e","#1e74d0")) +
  ylab("Simpson Diversity") +
  xlab("") +
  theme_minimal()

x <- ggpubr::ggarrange(p1, p2, p3,
                       ncol = 3,
                       common.legend = TRUE)
# T-test by year
# T-test
status1 <- 
  wilcox.test(diversity_rich$div[diversity_rich$year == "pan2020"], 
         diversity_rich$div[diversity_rich$year == "pan2021"])
rich_p <- t_status1$p.value

t_status2 <- 
  wilcox.test(diversity_shannon$div[diversity_rich$year == "pan2020"], 
         diversity_shannon$div[diversity_rich$year == "pan2021"])
shan_p <- t_status2$p.value

t_status3 <- 
  wilcox.test(diversity_simpson$div[diversity_rich$year == "pan2020"], 
         diversity_simpson$div[diversity_rich$year == "pan2021"])

simp_p <- t_status3$p.value

# Figure
# Facet by year, fill by status 
alpha_bar1 <- 
  diversity_rich %>%
  ggplot(aes(year, div)) + 
  geom_boxplot(aes(fill = year), alpha = 0.7) +
  geom_jitter(aes(color = year), alpha = 0.5) +
  theme_minimal() +
  stat_summary(fun = mean, geom = "point", shape = 24, size= 4, 
               color ="black", fill="white") 

alpha_bar2 <- 
  diversity_shannon %>% 
  ggplot(aes(year, div)) + 
  geom_boxplot(aes(fill = year), alpha = 0.7) +
  geom_jitter(aes(color = year), alpha = 0.5) +
  theme_minimal() +
  stat_summary(fun = mean, geom = "point", shape = 24, size= 4, 
               color ="black", fill="white") +
  annotate(geom = "text", 
           label = paste0("p-value: ", signif(t_status2$p.value, digits = 3)), 
           x = 1, y = 3.5) +
  theme_minimal()

alpha_bar3 <- 
  diversity_simpson %>%
  ggplot(aes(year, div)) + 
  geom_boxplot(aes(fill = year), alpha = 0.7) +
  geom_jitter(aes(color = year), alpha = 0.5) +
  theme_minimal() +
  stat_summary(fun = mean, geom = "point", shape = 24, size= 4, 
               color ="black", fill="white") 

x2 <- ggpubr::ggarrange(alpha_bar1, alpha_bar2, alpha_bar3,
                       ncol = 3,
                       labels = c("Richness",
                                  "Shannon",
                                  "Simpson"), common.legend = T)
x3 <- ggpubr::ggarrange(p2, alpha_bar2,
                        labels = c("Burn versus Unburn",
                                  "Fall versus Spring"))

habitat <- c("#ACC3A6", "#8E4677", "#F5D6BA", "#B6547D","#3F3F78")
habitat_div <- 
  diversity_shannon %>% 
  filter(!is.na(habitat)) %>%
  mutate(year = ifelse(year == "pan2020", "Fall 2020", "Spring 2021")) %>%
  ggplot(aes(year, div)) + 
  geom_boxplot(aes(fill = habitat), color = "black", alpha = 0.8, width = 0.5) +
  geom_jitter(alpha = 0.4) +
  stat_summary(aes(group = habitat),
               fun = median,
               geom = "line",
               alpha = 0.7,
               linewidth = 0.7) +
  facet_grid(habitat ~ status, as.table = TRUE) +
  scale_fill_manual(values = habitat) +
  theme_classic() +
  ylab("Shannon Diversity") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, 
                                   size = 8, 
                                   margin = margin(t = 14)),
        legend.position = "none")
ggsave("richness_habitat.png", width = 4, height = 5, units = "in")

```

Hypothesis 2: Burned and unburned sites will differ compositionally, using MOTUs
```{r ordination, adonis, betadis}
fire_color <- c( "#E47B3A", "#385653")
# Three or more OTUs per site
otu_counts_filt <- 
  df_filtered %>%
  group_by(sample_name) %>%
  filter(n_distinct(otu) >= 3) %>%
  ungroup() %>%
  group_by(sample_name, otu) %>%
  # Remove multiple rows
  filter(row_number() == 1) %>%
  # Use reads across all replicates
  summarise(otu_count = sum(rep_combined)) %>%
  ungroup()

# Create Hellinger, incidence matrix
otu_mat_filt <-
  otu_counts_filt %>%
  select(sample_name, otu, otu_count) %>%
  acast(sample_name ~ otu, fill = 0)

hell_mat_filt <- decostand(otu_mat_filt, method = "hellinger")
inc_mat_filt <- otu_mat_filt
inc_mat_filt[inc_mat_filt > 0] <- 1

dhell_mat <- hell_mat_filt
dinc_mat <- inc_mat_filt

# NMDS exploration - Hellinger distances
x <- .Random.seed
nmds <- metaMDS(dhell_mat, distance = "euclidean", 
                trymax = 100) # Only did 100 because it fails regardless with outliers
nmds_df <- 
  as_tibble(nmds$points, rownames = "sample_name") %>%
  left_join(df_filtered, by = "sample_name")

nmdseuc <- 
  nmds_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  geom_point(aes(color = status), size = 1) +
  geom_text(aes(label = sample_name, color = status), size = 1) +
  theme_minimal()

nmds <- metaMDS(dinc_mat, trymax = 100, distance = "jaccard")
nmds_df <- 
  as_tibble(nmds$points, rownames = "sample_name") %>%
  left_join(df_filtered, by = "sample_name")
nmdsjacc <- 
  nmds_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  geom_point(aes(color = status), size = 1) +
  geom_text(aes(label = sample_name, color = status), size = 1) +
  theme_minimal()

nmds_supp <- ggarrange(nmdseuc, nmdsjacc, labels = c("Hellinger Standardized", "Incidence"), common.legend = TRUE)

# Dataframe for permanova
permanova_df <-
  df_filtered %>%
  group_by(sample_name) %>%
  filter(row_number() == 1) %>%
  filter(sample_name %in% rownames(as.matrix(dhell_mat)))

# PERMANOVA
adonis_total_hell <- adonis2(dhell_mat ~ year*status*habitat*reserve,
                        method = "euclidean", data = permanova_df, 
                        permutations = 1000)
adonis_total_jac <- adonis2(dinc_mat ~ year*status*habitat*reserve,
                        method = "jaccard", data = permanova_df, 
                        permutations = 1000)
library(gt)
as.data.frame(adonis_total_hell) %>% rownames_to_column(var = "Predictor") %>% select(-Df) %>% gt()
as.data.frame(adonis_total_jac) %>% rownames_to_column(var = "Predictor") %>% select(-Df) %>% gt()

# PERMDISP
jac <- vegdist(dinc_mat, method = "jaccard")
euc <- vegdist(dhell_mat, method = "euclidean")

anova(betadisper(jac, permanova_df$status)) 
anova(betadisper(euc, permanova_df$status)) 

# Without outliers
outliers <- c("pan2021_BC_S01_Unburn",
              "pan2020_ML_O01_Burn")

mat_outlier_i <- 
  subset(dinc_mat, !rownames(dinc_mat) %in% outliers)
  
mat_outlier_h <- 
  subset(dhell_mat, !rownames(dhell_mat) %in% outliers)

nmds_outlier_jac <- metaMDS(mat_outlier_i, dist = "jaccard", trymax = 1000,
                        k = 3, noshare = TRUE)
nmds_outlier_euc <- metaMDS(mat_outlier_h, dist = "euclidean", trymax = 1000, 
                            k = 3, noshare = TRUE)

nmds_df <- 
  as_tibble(nmds_outlier_euc$points, rownames = "sample_name") %>%
  left_join(df_filtered, by = "sample_name")

fire_color <- c( "#E47B3A", "#385653")

nmds_jac <- 
  nmds_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  #geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(color = status), size = 2, alpha = 0.5) +
  stat_ellipse() +
  scale_color_manual(values = fire_color) +
  theme_minimal()

nmds_euc <- 
  nmds_df %>%
  ggplot(aes(x = MDS1, y = MDS2)) +
  #geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(color = status, shape = status), alpha = 0.3, size = 3) +
  stat_ellipse(aes(lty = status, color = status), alpha = 0.5, linewidth = 1) +
  scale_color_manual(values = fire_color) +
  theme_minimal() +
  geom_text(x = 1.8, y = 1.0, 
            label = paste("Stress = ", round(nmds_outlier_euc$stress, 3)), 
            size = 3, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.15, 
            label = "R2 = 0.013", 
            size = 4, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.3, 
            label = "F = 1.146", 
            size = 4, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.45, 
            label = "Pr = 0.158", 
            size = 4, alpha = 0.7) +
  labs(color = "Burn Status", 
       shape = "Burn Status",
       lty = "Burn Status")
ggsave("nmds_burn.png", width = 7, height = 5, units = "in")

# Dataframe for permanova
permanova_df <-
  df_filtered %>%
  group_by(sample_name) %>%
  filter(row_number() == 1) %>%
  filter(sample_name %in% rownames(as.matrix(mat_outlier_h)))

# PERMDISP

jac <- vegdist(mat_outlier_i, method = "jaccard")
euc <- vegdist(mat_outlier_h, method = "euclidean")

anova(betadisper(jac, permanova_df$status))
anova(betadisper(euc, permanova_df$status))

# Are there significant differences by status, habitat, reserve, year, etc?
adonis_total_hell <- adonis2(mat_outlier_h ~ year*status*habitat*reserve,
                        method = "euclidean", data = permanova_df, 
                        permutations = 1000)

adonis_total_inc <- adonis2(mat_outlier_i ~ year*status*habitat*reserve,
                        method = "euclidean", data = permanova_df, 
                        permutations = 1000)

nmds_year_p <- 
  nmds_df %>%
  mutate(year = ifelse(year == "pan2020", "Fall 2020", "Spring 2021")) %>%
  ggplot(aes(x = MDS1, y = MDS2)) +
  #geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(color = year, shape = year), alpha = 0.8, size = 3) +
  stat_ellipse(aes(color = year), alpha = 0.2, linewidth = 1) +
  scale_shape_manual(values = c(7, 8)) +
  scale_color_manual(values = c("#a384b3", "#41244f")) +
  theme_minimal() +
  geom_text(x = 1.8, y = 1.0, 
            label = paste("Stress = ", round(nmds_outlier_euc$stress, 3)), 
            size = 3, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.15, 
            label = "R2 = 0.034", 
            size = 4, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.3, 
            label = "F = 2.895", 
            size = 4, alpha = 0.7) +
  geom_text(x = 1.8, y = 1.45, 
            label = "Pr < 0.0001***", 
            size = 4, alpha = 0.7) +
  labs(color = "Season", 
       shape = "Season")
ggsave("nmds_season.png", width = 7, height = 5, units = "in")


  round(nmds_outlier_euc$stress, 3)
```

Distances
```{r}
distances <- 
  read.csv("fire_dist.csv") %>%
  rename(site1 = InputID,
         site2 = TargetID,
         distance = Distance) %>%
  mutate(site1 = gsub(" ", "", site1),
         site2 = gsub(" ", "", site2))
```

Beta diversity
```{r calculate beta diversity}
beta_bat <- beta(dhell_mat, func = "euclidean")
beta_total <- as.matrix(beta_bat$Btotal)
beta_repl <- as.matrix(beta_bat$Brepl)
beta_rich <- as.matrix(beta_bat$Brich)

# Eliminate double comparisons by making upper tri 0
beta_total[upper.tri(beta_total)] <- 0
beta_repl[upper.tri(beta_repl)] <- 0
beta_rich[upper.tri(beta_rich)] <- 0

beta_total <- 
  # Melt into DF
  melt(beta_total) %>%
  # Remove 0 values - self comparisons, or reverse comparisons
  filter(value != 0) %>%
  left_join(sites, by = c("Var1" = "sample_name")) %>%
  left_join(sites, by = c("Var2" = "sample_name")) %>%
  mutate(beta_type = "total")

beta_repl <- 
  # Melt into DF
  melt(beta_repl) %>%
  # Remove 0 values - self comparisons, or reverse comparisons
  filter(value != 0) %>%
  left_join(sites, by = c("Var1" = "sample_name")) %>%
  left_join(sites, by = c("Var2" = "sample_name")) %>%
  mutate(beta_type = "replacement")

beta_rich <- 
  # Melt into DF
  melt(beta_rich) %>%
  # Remove 0 values - self comparisons, or reverse comparisons
  filter(value != 0) %>%
  left_join(sites, by = c("Var1" = "sample_name")) %>%
  left_join(sites, by = c("Var2" = "sample_name")) %>%
  mutate(beta_type = "lossgain")

```

Beta against distances 
```{r}
beta_all <- rbind(beta_total,
                  beta_repl,
                  beta_rich) %>%
  rename(site1 = site_name.x,
         site2 = site_name.y) %>%
    mutate(comparison = case_when(
    status.x == status.y & status.x == "Unburn" ~ "Unburned",
    status.x == status.y & status.x == "Burn" ~ "Burned",
    TRUE ~ "Between"
  )) %>%
  left_join(distances, by = c("site1", "site2"))

within <- beta_all %>%
  filter(reserve.x == reserve.y
         & habitat.x == habitat.y 
         & year.x == year.y)

btwn <- beta_all %>%
    filter(reserve.x != reserve.y 
         & habitat.x == habitat.y 
         & year.x == year.y)

beta_dist_plot <- 
  within %>%
  #filter(beta_type != "total") %>%
  filter(comparison == "Between") %>%
  filter(distance < 6000) %>%
  mutate(year.x = ifelse(year.x == "pan2020", "Fall", "Spring")) %>%
  ggplot(aes(distance, value)) +
  geom_point(aes(color = year.x, shape = beta_type, size = beta_type), 
             alpha = 0.5) +
  geom_smooth(aes(lty = beta_type, color = year.x), method = "lm", 
              alpha = 0.1)  +
  facet_grid(~year.x) +
  xlab("Distance (km)") +
  ylab("Beta Diversity") +
  labs(color = "Season",
       shape = "Beta Components",
       lty = "Beta Components",
       size = "Beta Components") + 
  #scale_shape_manual(values = c(7, 8)) +
  scale_color_manual(values = c("#a384b3", "#41244f")) +
  scale_size_manual(values = c(2, 2.5, 1)) +
  scale_linetype_manual(values = c("dotted", "longdash", "solid")) +
  #(~comparison) +
  theme_minimal() 

df_distance <- within %>%
  #filter(beta_type != "total") %>%
  filter(comparison == "Between") %>%
  filter(distance < 6000) 

lm1 <- lm(df_distance$value[df_distance$beta_type == "total" & df_distance$year.x == "pan2020"] ~ df_distance$distance[df_distance$beta_type == "total" & df_distance$year.x == "pan2020"])
summary(lm1) #   0.031


lm2 <- lm(df_distance$value[df_distance$beta_type == "total" & df_distance$year.x == "pan2021"] ~ df_distance$distance[df_distance$beta_type == "total" & df_distance$year.x == "pan2021"])
summary(lm2) 
hist(lm1$residuals)


lm3 <- lm(df_distance$value[df_distance$beta_type == "lossgain" & df_distance$year.x == "pan2020"] ~ df_distance$distance[df_distance$beta_type == "lossgain" & df_distance$year.x == "pan2020"])
summary(lm3)
lm4 <- lm(df_distance$value[df_distance$beta_type == "lossgain" & df_distance$year.x == "pan2021"] ~ df_distance$distance[df_distance$beta_type == "lossgain" & df_distance$year.x == "pan2021"])
summary(lm4) # 0.051
hist(lm4$residuals)


lm5 <- lm(df_distance$value[df_distance$beta_type == "replacement" & df_distance$year.x == "pan2020"] ~ df_distance$distance[df_distance$beta_type == "replacement" & df_distance$year.x == "pan2020"])
summary(lm5) # 0.0345

lm6 <- lm(df_distance$value[df_distance$beta_type == "replacement" & df_distance$year.x == "pan2021"] ~ df_distance$distance[df_distance$beta_type == "replacement" & df_distance$year.x == "pan2021"])
summary(lm6)
```

Hypothesis 3: Burn / unburn will become more computationally similar from fall to spring - recovery (data combined)
```{r between site differences}

beta_narrowed <-
  beta_total %>%
  filter(reserve.x == reserve.y &
         habitat.x == habitat.y &
           year.x == year.y) %>%
  mutate(comparison = case_when(
    status.x == status.y & status.x == "Unburn" ~ "Unburned",
    status.x == status.y & status.x == "Burn" ~ "Burned",
    TRUE ~ "Between"))

beta_narrowed %>%
  ggplot(aes(comparison, value)) +
  geom_boxplot(aes(fill = year.x)) +
  geom_jitter(alpha = 0.1) +
  theme_minimal()
  
kruskal.test(beta_narrowed$value ~ beta_narrowed$comparison) #  p-value = 0.1365

# Summary
beta_narrowed %>%
  group_by(comparison) %>%
  summarise(median(value),
            mean(value))
year_box <- 
  beta_narrowed %>%
  ggplot(aes(as.factor(year.x), value)) +
  geom_boxplot(aes(fill = as.factor(year.x)), alpha = 0.5) +
  geom_jitter(aes(fill = as.factor(year.x)), alpha = 0.3) +
  facet_grid(~comparison) +
  stat_summary(fun = mean, geom = "point", shape = 24, size = 4, 
               color = "black", fill = "white") +
  theme_minimal() 

t.test(value ~ year.x, data = beta_narrowed)

```

Hypothesis 4: varies by habitat
```{r grassland nmds/permanova}
dinc_mat <- inc97_mat
dhell_mat <- hell97_mat

# pan2021_HT_G01_Burn, pan2021_AN_G05_Burn, pan2020_ML_G01_Burn
grass_out <- c( "pan2021_HT_G01_Burn", "pan2021_AN_G05_Burn", 
                "pan2021_AN_G03_Unburn")

## 2020 ##
# Matrix
inc_grass <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        sites$sample_name[sites$habitat == "Grassland" 
                                         & sites$year == "pan2020"])
inc_grass <- as.matrix(inc_grass[,colSums(inc_grass) != 0])
inc_grass <- subset(inc_grass, rownames(inc_grass) != "pan2020_ML_G01_Burn")

hell_grass <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        sites$sample_name[sites$habitat == "Grassland"
                                         & sites$year == "pan2020"])
hell_grass <- as.matrix(hell_grass[,colSums(hell_grass) != 0])
hell_grass <- subset(hell_grass, rownames(hell_grass) != "pan2020_ML_G01_Burn")

# Ordination
set.seed(1)
grass_mds_inc <- metaMDS(inc_grass, distance = "jaccard", trymax = 100, 
                         k = 2)
grass_mds_hell <- metaMDS(hell_grass, distance = "euclidean", trymax = 1000, 
                          k = 2)

# Plot
grass_df <- 
  as_tibble(grass_mds_hell$points, rownames = "sample_name") %>%
  left_join(sites, by = "sample_name") 

grass_2020 <- 
  grass_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  #geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.5) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
grass_dist <- vegdist(hell_grass)

# PERMANOVA / PERMDISP
perm_2020_grass <- adonis2(grass_dist~status, 
        sites[sites$sample_name %in% rownames(hell_grass),], 
        permutations = 1000)
ano_2020_grass <- anosim(hell_grass, distance = "euclidean",
       grouping = sites[sites$sample_name %in% rownames(hell_grass),]$status, 
       permutations = 1000)

dispr_2020_grass <- anova(betadisper(grass_dist, 
                 sites$status[sites$sample_name %in% rownames(hell_grass)]))

## 2021 ##
grass_out <- c( "pan2021_HT_G02_Unburn", "pan2021_HT_G04_Burn",
                "pan2021_AN_G03_Unburn")

# Matrix
inc_grass <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        sites$sample_name[sites$habitat == "Grassland" 
                                         & sites$year == "pan2021"])
inc_grass <- as.matrix(inc_grass[,colSums(inc_grass) != 0])
inc_grass <- subset(inc_grass, !rownames(inc_grass) %in% grass_out)

hell_grass <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        sites$sample_name[sites$habitat == "Grassland"
                                         & sites$year == "pan2021"])
hell_grass <- as.matrix(hell_grass[,colSums(hell_grass) != 0])
hell_grass <- subset(hell_grass, !rownames(hell_grass) %in% grass_out)

# Ordination
set.seed(1)
grass_mds_inc <- metaMDS(inc_grass, distance = "jaccard", trymax = 100, 
                         k = 2)
grass_mds_hell <- metaMDS(hell_grass, distance = "euclidean", trymax = 100, 
                          k = 2)

# Plot
grass_df <- 
  as_tibble(grass_mds_hell$points, rownames = "sample_name") %>%
  left_join(sites, by = "sample_name") 

grass_2021 <- 
  grass_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  # geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.5) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
grass_dist <- vegdist(hell_grass)

# PERMANOVA / PERMDISP
perm_2021_grass <- adonis2(grass_dist~status, 
        sites[sites$sample_name %in% rownames(hell_grass),], 
        permutations = 1000)
ano_2021_Grass <- anosim(hell_grass, distance = "euclidean",
       grouping = sites[sites$sample_name %in% rownames(hell_grass),]$status, 
       permutations = 1000)
dispr_2021_grass <- anova(betadisper(grass_dist, 
                 sites$status[sites$sample_name %in% rownames(hell_grass)]))


# Richness

diversity_shannon %>% 
  filter(habitat == "Grassland") %>%
  ggplot(aes(status, div)) +
  geom_boxplot(aes(color = year)) +
  geom_jitter(alpha = 0.3) 

# Beta components

beta_all %>%
  filter(habitat.x == "Grassland" & habitat.y == "Grassland"
         & year.x == year.y & reserve.x == reserve.y) %>%
  mutate(comp = case_when(
    status.x == status.y & status.x == "Unburn" ~ "Unburned",
    status.x == status.y & status.x == "Burn" ~ "Burned",
    status.x != status.y ~ "Between"
  )) %>%
  filter(comp == "Between") %>%
  ggplot(aes(distance, value)) +
  geom_point(aes(color = beta_type)) +
  geom_smooth(method = "lm") +
  facet_grid(~year.x)

```

```{r scrub nmds/permanova}
site <- sites

# pan2021_HT_G01_Burn, pan2021_AN_G05_Burn, pan2020_ML_G01_Burn
scrub_out <- c( "pan2020_AN_S01_Burn")

## 2020 ##
# Matrix
inc_scrub <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        site$sample_name[site$habitat == "Scrub" 
                                         & site$year == "pan2020"])
inc_scrub <- as.matrix(inc_grass[,colSums(inc_grass) != 0])
inc_scrub <- subset(inc_scrub, !rownames(inc_scrub) %in% scrub_out)

hell_scrub <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        site$sample_name[site$habitat == "Scrub"
                                         & site$year == "pan2020"])
hell_scrub <- as.matrix(hell_scrub[,colSums(hell_scrub) != 0])
hell_scrub <- subset(hell_scrub, !rownames(hell_scrub) %in% scrub_out)

# Ordination
set.seed(1)
scrub_mds_inc <- metaMDS(inc_scrub, distance = "jaccard", trymax = 1000, 
                         k = 2, noshare = TRUE)
scrub_mds_hell <- metaMDS(hell_scrub, distance = "euclidean", trymax = 1000, 
                          k = 2, noshare = TRUE)

# Plot
scrub_df <- 
  as_tibble(scrub_mds_hell$points, rownames = "sample_name") %>%
  left_join(site, by = "sample_name") 

scrub_2020 <- 
  scrub_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  # geom_text(aes(label = sample_name), size = 1) +
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.4) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
scrub_dist <- vegdist(hell_scrub)

# PERMANOVA / PERMDISP
perm_2020_scrub <- adonis2(scrub_dist~status, 
        site[site$sample_name %in% rownames(hell_scrub),], 
        permutations = 1000)
ano_2020 <- anosim(hell_scrub, distance = "euclidean",
       grouping = site[site$sample_name %in% rownames(hell_scrub),]$status, 
       permutations = 1000)
dispr_2020 <- anova(betadisper(scrub_dist, 
                 site$status[site$sample_name %in% rownames(hell_scrub)]))

## 2021 ##
scrub_out <- c("pan2021_BC_S01_Unburn")

# Matrix
inc_scrub <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        site$sample_name[site$habitat == "Scrub" 
                                         & site$year == "pan2021"])
inc_scrub <- as.matrix(inc_grass[,colSums(inc_grass) != 0])
inc_scrub <- subset(inc_scrub, !rownames(inc_scrub) %in% scrub_out)

hell_scrub <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        site$sample_name[site$habitat == "Scrub"
                                         & site$year == "pan2021"])
hell_scrub <- as.matrix(hell_scrub[,colSums(hell_scrub) != 0])
hell_scrub <- subset(hell_scrub, !rownames(hell_scrub) %in% scrub_out)

# Ordination
set.seed(1)
scrub_mds_inc <- metaMDS(inc_scrub, distance = "jaccard", trymax = 1000, 
                         k = 2, noshare = TRUE)
scrub_mds_hell <- metaMDS(hell_scrub, distance = "euclidean", trymax = 1000, 
                          k = 2, noshare = TRUE)

# Plot
scrub_df <- 
  as_tibble(scrub_mds_hell$points, rownames = "sample_name") %>%
  left_join(site, by = "sample_name") 

scrub_2021 <- 
  scrub_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.4) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
scrub_dist <- vegdist(hell_scrub)

# PERMANOVA / PERMDISP
perm_2021_scrub <- adonis2(scrub_dist~status, 
        site[site$sample_name %in% rownames(hell_scrub),], 
        permutations = 1000)
ano_2021 <- anosim(hell_scrub, distance = "euclidean",
       grouping = site[site$sample_name %in% rownames(hell_scrub),]$status, 
       permutations = 1000)
dispr_2021 <- anova(betadisper(scrub_dist, 
                 site$status[site$sample_name %in% rownames(hell_scrub)]))

scrub <- ggpubr::ggarrange(scrub_2020, scrub_2021, 
                           labels = c("Fall 2020", "Spring 2021"),
                           common.legend = TRUE)
```

```{r oak nmds/permanova}
site <- sites

# pan2021_HT_G01_Burn, pan2021_AN_G05_Burn, pan2020_ML_G01_Burn
oak_out <- c( "pan2020_AN_S01_Burn", "pan2020_AN_S01_Unburn")

## 2020 ##
# Matrix
inc_oak <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        site$sample_name[site$habitat == "Oak Woodland" 
                                         & site$year == "pan2020"])
inc_oak <- as.matrix(inc_oak[,colSums(inc_oak) != 0])
# inc_oak <- subset(inc_oak, !rownames(inc_oak) %in% oak_out)

hell_oak <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        site$sample_name[site$habitat == "Oak Woodland"
                                         & site$year == "pan2020"])
hell_oak <- as.matrix(hell_oak[,colSums(hell_oak) != 0])
hell_oak <- subset(hell_oak, !rownames(hell_oak) %in% oak_out)

# Ordination
set.seed(1)
oak_mds_inc <- metaMDS(inc_oak, distance = "jaccard", trymax = 100, 
                         k = 2)
oak_mds_hell <- metaMDS(hell_oak, distance = "euclidean", trymax = 100, 
                          k = 2)

# Plot
oak_df <- 
  as_tibble(oak_mds_hell$points, rownames = "sample_name") %>%
  left_join(site, by = "sample_name") 

oak_2020 <- 
  oak_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.4) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
oak_dist <- vegdist(hell_oak)

# PERMANOVA / PERMDISP
perm_2020 <- adonis2(oak_dist~status, 
        site[site$sample_name %in% rownames(hell_oak),], 
        permutations = 1000)
dispr_2020 <- anova(betadisper(oak_dist, 
                 site$status[site$sample_name %in% rownames(hell_oak)]))

## 2021 ##
# Matrix
inc_oak <- subset(dinc_mat, rownames(dinc_mat) %in% 
                        site$sample_name[site$habitat == "Oak Woodland" 
                                         & site$year == "pan2021"])
inc_oak <- as.matrix(inc_oak[,colSums(inc_oak) != 0])
# inc_oak <- subset(inc_oak, !rownames(inc_oak) %in% oak_out)

hell_oak <- subset(dhell_mat, rownames(dhell_mat) %in% 
                        site$sample_name[site$habitat == "Oak Woodland"
                                         & site$year == "pan2021"])
hell_oak <- as.matrix(hell_oak[,colSums(hell_oak) != 0])
hell_oak <- subset(hell_oak, !rownames(hell_oak) %in% "pan2021_BO_O03_Burn")

# Ordination
set.seed(1)
oak_mds_inc <- metaMDS(inc_oak, distance = "jaccard", trymax = 100, 
                         k = 2)
oak_mds_hell <- metaMDS(hell_oak, distance = "euclidean", trymax = 100, 
                          k = 2, noshare = TRUE)

# Plot
oak_df <- 
  as_tibble(oak_mds_hell$points, rownames = "sample_name") %>%
  left_join(site, by = "sample_name") 

oak_2021 <- 
  oak_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = status)) + 
  geom_point(aes(shape = status), size = 2) +
  stat_ellipse(aes(lty = status), alpha = 0.4) +
  scale_color_manual(values = fire_color) +
  theme_minimal() 

# Distance matrix
oak_dist <- vegdist(hell_oak)

# PERMANOVA / PERMDISP
perm_2021 <- adonis2(oak_dist~status, 
        site[site$sample_name %in% rownames(hell_oak),], 
        permutations = 1000)
dispr_2021 <- anova(betadisper(oak_dist, 
                 site$status[site$sample_name %in% rownames(hell_oak)]))

oak_nmds <- ggpubr::ggarrange(oak_2020, oak_2021, 
                  common.legend = TRUE, 
                  labels = c("Fall 2020", "Spring 2021"))


```

```{r total NMDS}
grass_nmds <- ggpubr::ggarrange(grass_2020, grass_2021, 
                                oak_2020, oak_2021,
                                scrub_2020, scrub_2021,
                                ncol = 2, nrow = 3,
                                common.legend = TRUE)

```

```{r beta diversity by habitat}
hab_beta <-
  beta_narrowed %>%
  #filter(comparison == "Between") %>%
  ggplot(aes(comparison, value)) +
  geom_boxplot(aes(fill = habitat.x), alpha = 0.8) +
  geom_jitter(aes(shape = comparison), alpha = 0.3) +
  facet_grid(~habitat.x) +
  scale_color_manual(values = fire_color) +
  scale_fill_manual(values = habitat) +
  labs(fill = "Habitat", 
       shape = "Burn Status") +
  xlab("") +
  ylab("Beta Diversity") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6, angle = 45),
        legend.key.size = unit(0.75, "lines"),
        legend.text = element_text(size = 6))
ggsave("hab_beta.png", width = 6, height = 3.5, units = "in")

beta_narrowed %>%
  group_by(comparison, habitat.x) %>%
  summarise(median(value),
            mean(value))


hab_beta_between <- beta_narrowed %>% filter(comparison == "Between")

kruskal.test(hab_beta_between$value ~ hab_beta_between$habitat.x)

kruskal.test(hab_beta_between$value ~ hab_beta_between$year.x)

kruskal.test(hab_beta_between$value[hab_beta_between$habitat.x == "Scrub"] ~ hab_beta_between$year.x[hab_beta_between$habitat.x == "Scrub"]) # 0.059

kruskal.test(hab_beta_between$value[hab_beta_between$habitat.x == "Grassland"] ~ hab_beta_between$year.x[hab_beta_between$habitat.x == "Grassland"]) # 0.07495

kruskal.test(hab_beta_between$value[hab_beta_between$habitat.x == "Oak Woodland"] ~ hab_beta_between$year.x[hab_beta_between$habitat.x == "Oak Woodland"]) # 0.07495

beta_reserve_diff <- 
  beta_total %>%
  mutate(comparison = case_when(
    status.x == status.y & status.x == "Unburn" ~ "Unburned",
    status.x == status.y & status.x == "Burn" ~ "Burned",
    TRUE ~ "Between")) %>%
  filter(beta_type == "total") %>%
  filter(habitat.x == habitat.y &
           reserve.x != reserve.y) 

beta_reserve_diff %>%
  ggplot(aes(comparison, value)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2) +
  facet_grid(~habitat.x)

kruskal.test(beta_reserve_diff$value[beta_reserve_diff$habitat.x == "Scrub"] ~ beta_reserve_diff$comparison[beta_reserve_diff$habitat.x == "Scrub"])

kruskal.test(beta_reserve_diff$value[beta_reserve_diff$habitat.x == "Grassland"] ~ beta_reserve_diff$comparison[beta_reserve_diff$habitat.x == "Grassland"])

kruskal.test(beta_reserve_diff$value[beta_reserve_diff$habitat.x == "Oak Woodland"] ~ beta_reserve_diff$comparison[beta_reserve_diff$habitat.x == "Oak Woodland"])


beta_all %>%
  filter(habitat.x == habitat.y,
         reserve.x == reserve.y) %>%
  mutate(comparison = case_when(
    status.x == status.y & status.x == "Unburn" ~ "Unburned",
    status.x == status.y & status.x == "Burn" ~ "Burned",
    TRUE ~ "Between")) %>%
  #filter(beta_type == "total") %>%
  filter(distance < 6000) %>%
  ggplot(aes(distance, value)) +
  geom_point(aes(color = beta_type)) +
  geom_smooth(aes(color = beta_type), method = "lm", alpha = 0.1) +
  facet_grid(comparison~habitat.x) +
  theme_minimal()



```
